---
title: "Analysis of Fatal Police Shootings in the United States (2015-2022)"
output: html_document
date: "2023-10-05"
---


# Introduction

The dataset *all_police_shootings* used in this analysis is a record of police shootings in the United States from 2015 to 2022 time period. Police shootings are a topic of immense societal concern and have sparked nationwide debates and discussions. These incidents often raise questions about the use of force, racial disparities, and the broader issues surrounding public safety.
In this analysis, we are interested in investigating the factors that affect the likelihood of being involved in a police shooting


# Preliminary work
### Installing Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#getwd()

library("tidyverse")
library("knitr")
library("ggplot2")
library("plotly")
```
### Installing .csv file into R
```{r}
all_police_shootings <- read.csv("data/fatal-police-shootings-data.csv")
```
# Basic Data Analysis
### Summary Statistics
From the first few rows of the dataset, we gain a basic understanding of what information the dataset contains: the name of the person shot, date of incident, manner of death, type of arm equipped, age, gender and race of the person, as well as the geographic information of the incident, represented by city name, state name and longitude and latitude value. The dataset also contains some boolean/binary values that details if the person has signs of mental illness, if the person is fleeing and if the police wears body_camera. 

```{r}
print(head(all_police_shootings))
```

From the summary of data, we can see the age of the people shot ranges from 2 to 92, averaging to around 37 years old, and in most cases, there is no sign of mental illness with the police not wearing body camera. 

```{r}
summary(all_police_shootings)

```

# Data Exploration
### Count Number of Subjects
In the dataset, there are a total of 8002 observations and 17 variables.
```{r}
dim(all_police_shootings)
```
### Structure and Missing Data
Gain information about the data type of each variable as well as a preview of some of their possible values.
```{r}
str(all_police_shootings)
sapply(all_police_shootings, class)
```
Most variables have no missing values, while there are 840 observations missing longitude and latitude information and 503 entries missing age information.
```{r}
colSums(is.na(all_police_shootings))
```
## Frequency Tables
From the frequency tables, we can observe that there is a diverse range of weapon types recorded in the dataset. Some observations indicate the presence of a single weapon, while others suggest combinations of multiple weapons used in the incidents. Additionally, it's noteworthy that there are instances where individuals involved in the incidents were unarmed.

```{r}
manner_of_death_freq_table <- table(all_police_shootings$manner_of_death)
armed_freq_table <- table(all_police_shootings$armed)
age_freq_table <- table(all_police_shootings$age)
gender_freq_table <- table(all_police_shootings$gender)
race_freq_table <- table(all_police_shootings$race)
gender_freq_table <- table(all_police_shootings$gender)
state_freq_table <- table(all_police_shootings$state)
mental_illness_freq_table <- table(all_police_shootings$signs_of_mental_illness)
flee_freq_table <- table(all_police_shootings$flee)
body_camera_freq_table <- table(all_police_shootings$body_camera)

print(manner_of_death_freq_table)
print(armed_freq_table)
print(age_freq_table)

```

This gender distribution highlights a significant disparity, with a substantially higher number of male individuals involved in the recorded incidents compared to their female counterparts.
```{r}
print(gender_freq_table)
```

The largest group is white individuals, followed by black and Hispanic individuals. Apart from "others", Asian is the smallest group. 
```{r}
print(race_freq_table)
print(state_freq_table)
print(mental_illness_freq_table)
print(flee_freq_table)
print(body_camera_freq_table)
```


```{r}
# extract date and cast to type Date
all_police_shootings$date <- all_police_shootings$date %>% as.Date(format = "%m/%d/%Y")
all_police_shootings$day <- all_police_shootings$date %>% as.Date(format = "%m/%d/%Y") %>% format(format = "%d")
all_police_shootings$month <- all_police_shootings$date %>% as.Date(format = "%m/%d/%Y") %>% format(format = "%m")
all_police_shootings$year <- all_police_shootings$date %>% as.Date(format = "%m/%d/%Y") %>% format(format = "%Y")
```

# Visualizations
The line plot provides an overview of the trend in the population involved in police shootings spanning from 2015 to 2022. There is a dip in the year 2016 and 2022; however, the overall trend reveals a steady increase in incidents over this period, indicating a rising occurrence of such events.

```{r}
# how the number of incidents per year changes
counts_by_year <- table(all_police_shootings$year)
counts_by_year_df <- data.frame(Year = as.integer(names(counts_by_year)), Incidents = counts_by_year)

counts_by_year_df %>% ggplot(aes(x = Year, y = Incidents.Freq)) + geom_line() + geom_point() +
  labs(title = "Incidents by Year", x = "Year", y = "Count")
```

The age distribution of people involved in incidents follows a bell shape that concentrates around 20-40 years old. This distribution may be attributed to the combination of factors such as stronger physical capabilities and increased social interactions that individuals typically experience during this phase of their lives.  
```{r warning=FALSE}
all_police_shootings %>% filter(!is.na(age)) %>% 
  ggplot(aes(x = age)) + geom_histogram(bins = 20, fill = 'grey', color = 'black') + 
  ggtitle ("Age distribution of people involved in incidents")
```

As evident from the numerical data, this bar plot provides a clear visual representation of the gender disparity within the dataset. This disparity may be attributed to a higher probability of aggressive behavior during police encounters among males.
```{r warning=FALSE}
all_police_shootings %>% filter(gender != "") %>% ggplot(aes(x = gender)) + geom_bar() + 
  xlab("Gender") + ylab("Count") + ggtitle("Police Shooting by Gender")
```

Though there are substantially more males involved than females, within each gender group the age distribution has a similar pattern.
```{r warning=FALSE}
all_police_shootings %>% filter(gender != "") %>% ggplot(aes(x = gender, y = age)) + geom_boxplot() + 
  xlab("Gender") + ylab("Age") + ggtitle("Age Distribution by Gender")
```

As we noted, there are substantially more White people involved than Asian and Native American. This might also be attributed to the demographic where Whites make up about 75% of population in US. However, while Black only make up around 15% of US population, the number of Black involved in police shootings is half of that of the White, the reason of which might worth investigating.
```{r warning=FALSE}
all_police_shootings %>% filter(race != "") %>% ggplot(aes(x = race)) + geom_bar() + 
    labs(title = "Composition of Races in Police Shootings") +
    xlab("Race")
```

The age distribution within each racial group is mostly similar with only modest differences. Notably, the White racial group exhibits the widest span of age groups.
```{r warning=FALSE}
all_police_shootings %>% filter(race != "") %>% ggplot(aes(x = race, y = age)) + geom_boxplot() + 
  xlab("Race") + ylab("Age") + ggtitle("Age Distribution by Race")
```

From the scatter plot, we can see how  people involved are geographically distributed by gender. Regardless of gender, most incidents concentrate to the northwest region and border/coastal region. More females in the north are involved than in the south.
```{r warning=FALSE}
# Note the points outside of main region
all_police_shootings %>% filter(!is.na(longitude) & !is.na(latitude)) %>%
  ggplot(aes(x = latitude, y = longitude, color = gender)) + geom_point(alpha=0.6, size = 0.7) + 
  scale_color_manual(values = c("#55AD89", "#EF6F6A", "#1170AA")) + ggtitle("Geographic Distribution by Gender")
```


# Possible Research Questions

The dataset presents numerous avenues for research inquiries. Broadly, we can explore whether demographic factors are associated with the likelihood of being involved in a police shooting. Additionally, investigating the influence of mental illness, the presence of weapons, or the use of body cameras in these incidents can yield valuable insights. Temporal analysis can uncover trends over time, shedding light on shifts and patterns in these incidents.

To delve into specific areas, given the existing research that indicates a disproportionate victimization of Black individuals, it is essential to investigate the underlying factors contributing to this phenomenon. Furthermore, a focused research endeavor can center on the unique group of 'unarmed' individuals, aiming to understand the circumstances and dynamics surrounding their involvement in police shootings.

```{r include=FALSE}
## Data Cleaning

# -------------------------
###  Ideas & draft
# unarmed people's age, gender and race
all_police_shootings %>% filter(armed=="unarmed") %>%
  ggplot(aes(x = body_camera, color = race)) + geom_bar() # + scale_color_manual(values = c("#55AD89", "#EF6F6A", "#1170AA", "#FFD700", "#8566AA", "#FFA500", "#6A5ACD"))


```

```{r include=FALSE}
# NOTES
# how to organize the draft
# 1. By code function (tables, plots...)
# 2. By variables (age, gender, ...)
# need to get rid of empty values in race 
# categorize weapons?
# how to make race plots not messy
# all_police_shootings %>% filter(race != "")
# plot: racial bias by geography --> too many categories

unique(all_police_shootings$armed)

library(shiny)

ui <- fluidPage(
  titlePanel("Racial Distribution in Police Shootings Over Time"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("year", "Select Year", min = min(all_police_shootings$date), 
                  max = max(all_police_shootings$date), value = min(all_police_shootings$date))
    ),
    mainPanel(
      plotlyOutput("plot")
    )
  )
)


server <- function(input, output) {
  output$plot <- renderPlotly({
    filtered_data <- subset(all_police_shootings, year == input$date)
    
    plot_ly(
      data = filtered_data,
      x = ~date,
      y = ~frequency(),
      color = ~date,
      type = "bar"
    ) %>%
      layout(
        title = paste("Racial Distribution in", input$year, "Police Shootings"),
        xaxis = list(title = "Race"),
        yaxis = list(title = "Count"),
        barmode = "stack"
      )
  })
}

# shinyApp(ui, server)

```


